---
layout: post
title: "Encryption"
author: "Nanu Frechen"
excerpt: "How does Encryption work?"
category: howto
language: English
---
```{r settings, results='hide', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, cache=FALSE}

load=FALSE

library(knitr)

knit_hooks$set(small.mar = function(before, options, envir) {
  if (before) {
			par(mar=c(4,4,0.1,0.1), las=1, pch=20)
  	}
})

opts_chunk$set(message = F, error = F, warning = F, cache = TRUE, autodep=T, fig.width=7, fig.height=5, dev="svg", dpi=100, tidy=T, small.mar=T, tidy.opts=list(width.cutoff=60))

library(rgl)
library(pander)
panderOptions('table.style', 'rmarkdown')
panderOptions('table.split.table', 200)
```


* auto-gen TOC:
{:toc}

------------------------------------

Have you ever wondered how encryption works? It seems to be very advanced mathematics. Encryption algorithms are very complex and hard to understand. But is it possible to understand the basic principles, or even calculate it yourself? Here we will show some basic algorithms to do symmetric and asymmetric encryption that will help you to understand how it works.


# Symmetric Encryption

Symmetric Encryption means that you use the same key for encrypting and decrypting the message. This leaves you whith the problem of how to inform your recipient about what the key is.

Let's start with a very simple case, the old "exchange every letter for another". So we start with a message like this:

```{r message}
mes <- "Attack next Friday early in the Morning"
```

To keep it simple we translate the string to uppercase letters and replace the spacing between words with "_":

```{r uppercase}
mes <- toupper(mes)
mes <- gsub(" ", "_", mes)
```
So the message becomes "`r mes`".


Finally we split the string into a vector of single letters:
```{r split}
mes <- strsplit(mes, "")[[1]]
mes

```

## Define the key

To produce our key we create a table with all the letters from the alphabet and assign each with a new letter. You could choose the assignment yourself (or according to the order of first appearance in a book both you and your recipient possess). Here we just make a random shuffle of the letters with the function `sample()`:
```{r key}
LETTERSplus <- c(LETTERS, "_", ".", ",", "-")
key <- sample(LETTERSplus)
names(key) <- LETTERSplus
```

We end up with a table like this:
```{r print-key, echo=F}
pander(key, split.table=250)
```

which is our key. To encrypt we exchange each letter from the first line with the corresponding letter form the second line. To decrypt we exchange each letter from the bottom line with the corresponding letter from the first line.

## Encrypt the message
So our message would translate like this:
```{r print-encrypted-message, echo=F, cache=F}
pander(key[mes])
```

We achieve this in R by executing:
```{r save-mes_enc, results="hide"}
mes_enc <- key[mes]
```

To get a string we paste the letters in the vector together like this:
```{r mes_enc_pasted, eval=T}
paste(mes_enc, collapse="")
```

# Can we crack the code?

The letters in our encrypted message are not random. There is some structure in it that becomes obvious when look at the frequency the letters occur:

```{r eval=F}
table(mes_enc)
```

```{r, echo=F, cache=F}
pander(sort(table(mes_enc), decreasing=T))
```

Some letters are missing.

This becomes more obvious if we have a longer message:
```{r new-message, tidy=T}
new_mes <- 'A German Enigma operator would be given a plaintext message to encrypt. For each letter typed in, a lamp indicated a different letter according to a pseudo-random substitution, based upon the wiring of the machine. The letter indicated by the lamp would be recorded as the enciphered substitution. The action of pressing a key also moved the rotor so that the next key press used a different electrical pathway, and thus a different substitution would occur. For each key press there was rotation of at least the right hand rotor, giving a different substitution alphabet. This continued for each letter in the message until the message was completed and a series of substitutions, each different from the others, had occurred to create a cyphertext from the plaintext. The cyphertext would then be transmitted as normal to an operator of another Enigma machine. This operator would key in the cyphertext and - as long as all the settings of the deciphering machine were identical to those of the enciphering machine - for every key press the reverse substitution would occur and the plaintext message would emerge.'
```
(The text was taken from the [wikipedia page about the Enigma machine](https://en.wikipedia.org/wiki/Enigma_machine)).

We use the same procedure to encrypt this message:

```{r encrypt-new_mes}
new_mes <- toupper(new_mes)
new_mes <- gsub(" ", "_", new_mes)
new_mes <- strsplit(new_mes, "")[[1]]
new_mes_enc <- key[new_mes]
```

and end up with this:
```{r}
paste(new_mes_enc, collapse = "")
```

If we look at the letter frequency in this message we can draw this barplot:
```{r}
new_mes_freq <- sort(table(new_mes_enc), decreasing = T)
```

```{r, echo=F}
par(mgp=c(2,1,-.8))
barplot(new_mes_freq/length(new_mes)*100, xlab="letter", ylab="frequency [%]", cex.names=0.6)
```

## compare to english word frequency

```{r download-text, results="hide"}
# download the book
text <- readLines(
  "https://archive.org/download/adventuresoftoms00twai2/adventuresoftoms00twai2_djvu.txt")
# paste all lines together
text <- paste(text, collapse = "")
# convert everything to uppercase:
text <- toupper(text)
# remove all non-alphabetical letters
text <- strsplit(gsub("[^-.,[:alpha:] ]", "", text), " +")[[1]]
# paste it back together
text <- paste(text, collapse=" ")
# replace spaces by "_"
text <- gsub(" ", "_", text)
# split to vector of single letters:
text <- strsplit(text, "")[[1]]
# show the letter frequency:
sort(table(text), decreasing = T)/length(text)*100
```

```{r sorted-word-frequency, echo=F, cache=F, results='asis'}
text_freq <- sort(table(text), decreasing=TRUE)
pandoc.table(text_freq, split.tables=90)
```

```{r, echo=F}
par(mgp=c(2,1,-.8))
barplot(text_freq/length(text)*100, xlab="letter", ylab="frequency [%]", cex.names=0.6)
```


```{r, cache=F}
crack <- names(text_freq)
names(crack) <- c(names(new_mes_freq), NA, NA)
pander(sort(crack))

```


```{r, cache=F}
pander(key)

```


```{r}
paste(crack[new_mes_enc], collapse="")
```


```{r more-stats, fig.height=1.5, eval=F}
#new_mes <- paste(new_mes, collapse="")
test_text <- paste(text, collapse="")
par(mar=c(2,4,0.2,0.1), las=1, mgp=c(3,1,-0.6))
for(len in 1:5){
  test <- substring(test_text, 1:(nchar(test_text)-len+1), len:nchar(test_text))
  
  barplot(head(sort(table(test), decreasing = T)/length(test),10), cex.names=0.7, ylab="frequency [%]")
  
}
```

```{r, eval=F, echo=F}
for(i in 1:10){
  crack <- c(sample(names(sort(new_mes_freq)), prob=sort(c(new_mes_freq))^4), NA, NA, NA)
  print(t(data.frame(crack=crack, key=key[names(text_freq)], match=crack==key[names(text_freq)])))
  
}
```


# Asymmetric Encryption

```{r message2}
mes <- "Meet next sunday at the old building."

```


This time we don't replace our message by other characters, but by numbers in the following pattern:

$$remainder(\frac{plain^{key}}{33})=enc$$

Let us define this as a function:
```{r}
encrypt <- function(text, key, base=33){
  text^key %% base
}
```


Before we can use this function we have to translate our text tu numbers. One way is to convert all characters to "raw" which means into the bytes the computer stores this character:
```{r mes_num, dependson="message2"}
mes_num <- charToRaw(mes)
mes_num
```

This is still in hexadecimal format, which we can't use for calculations. So we convert to numeric:
```{r mes_num2, dependson="mes_num"}
mes_num <- as.integer(mes_num)
mes_num
```

At this point we still can easily convert it back:
```{r}
rawToChar(as.raw(mes_num))
```


Now we apply the encryption:
```{r sym-encrypt, dependson="mes_num2"}
mes_num_enc <- encrypt(text = mes_num, key = 3, base = 121)

```
Note how the numeric vector changed:
```{r}
mes_num_enc
```

We can try to convert it back to characters, but this gives us only a lot of rubbish:
```{r, eval=F}
rawToChar(as.raw(mes_num_enc))
```

The difference is now: we can't use the same key to decrypt the message again. The remainder calculation (done with `%%` in R) is a one-way calculation as shall be illustrated by the follwing:
```{r}
input <- 1:15
remainder <- input %% 3
```
`r pander(t(data.frame(input, remainder)), caption="remainder of the devision by 3:")`
You can see that the backwards calculation is not possible, because there are multiple remainders with the same numerical value produced by different numbers.

But due to a fancy mathematical condition the following is possible:
```{r basic-encrypt, eval=F}
base=33
mes_num_enc <- encrypt(text = mes_num, key = 3, base = base)

mes_num_dec <- encrypt(mes_num_enc, key=7, base=base)
mes_num_dec
mes_num
```

```{r, eval=F}
rawToChar(as.raw(mes_num_dec))
```


## Asymmetric encryption with small prime numbers
Let us start with a simple example:
```{r small-primes, eval=F}
while(!require(numbers)) install.packages("numbers")

q = 3
p = 11
(n = p * q)

(fn <- n - (p + q - 1))
# 1 < e < φ(n)
# gcd(e, φ(n)) = 1
e=3
GCD(e, fn)

m = 1:32 # message
(c = m^e %% n)
(dec <- c^7 %% n)
all(dec == m)
```


## Asymmetric encryption with large prime numbers
If you want to use large prime numbers to do the RSA encryption you need a modulo function that can handle large numbers. The default R `%%` modulo throws a warning on large numbers and returns nonsense.

To calculate with large numbers we use the package `gmp`. This package provides Multiple Precision Arithmetic using the C library GMP (GNU Multiple Precision Arithmetic).
```{r large-numbers}
while(!require(gmp)) install.packages("gmp")

two <- as.bigz(2)
p1 <- two^107 -1 
isprime(p1)
p1

p2 <- two^127 -1 
isprime(p2)
p2

p1 = 61
p2 = 53
(n = p1 * p2)
(t <- (p1-1)*(p2-1))
e = as.bigz(17)
(d = inv.bigz(e, t)) # 2753

m <- as.bigz(mes_num)
(c <- mod.bigz(m^e, n))
mes_num_dec <- as.integer(mod.bigz(pow.bigz(c, d), n))

identical(mes_num, mes_num_dec)
rawToChar(as.raw(mes_num_dec))
```


# Checksums
```{r eval=F}
library(digest)
library(PKI)
```


# Passwords
p
