---
layout: post
title: Critical transitions
excerpt: When stabile systems transition into instability
category: philosophical
language: English
---
```{r settings, results='hide', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, cache=FALSE}

load=FALSE

library(knitr)

knit_hooks$set(small.mar = function(before, options, envir) {
  if (before) {
			par(mar=c(4,4,.1,4), pch=20, mgp=c(2.5,1,0), las=1)
  	}
})

opts_chunk$set(message = F, error = F, warning = F, cache = TRUE, autodep=T, fig.width=7, fig.height=3, dev="svg", dpi=100, tidy=T, small.mar=T, tidy.opts=list(width.cutoff=60))

library(pander)
panderOptions('table.style', 'rmarkdown')
panderOptions('table.split.table', 200)
```

What is a stable system? A stable system is one that, if you push it out of balance, it will find it's balance again. Maybe it will take a while, maybe it will oscillate around the balance point, but finally it will be back in it's stable spot. A stable system must not be a static one. If it is under constant disturbance it will constantly be out of balance - but it will allways strive back to the balance point.

While a stable system is like a sack hanging on a rope or a ball in a bowl, an unstable system is like a pencil standing on it's tip. It is nearly impossible to stay in balance that way. Even the slightest disturbance will tip it over. If it is slightly out of it's balance spot it will accellerate more and more away from it. 

If we bring our imagination away from objects standing or our table (or not) there are systems that transition between a relatively stable state to a very unstable state. And we can observer this transition in how they react to disturbances. 

* tips into an alternative state

[Why should we expect early warnings before critical transitions?](http://www.early-warning-signals.org/theory/why-should-we-expect-early-warning/)

[earlywarnings](https://cran.r-project.org/web/packages/earlywarnings/index.html)

# The early warning signals toolbox
> **Early Warning Signals Toolbox for Detecting Critical Transitions in Timeseries**
> The Early-Warning-Signals Toolbox provides methods for estimating statistical changes in timeseries that can be used for identifying nearby critical transitions. Based on Dakos et al (2012) Methods for Detecting Early Warnings of Critical Transitions in Time Series Illustrated Using Simulated Ecological Data.



```{r install, eval=F}
install.packages("earlywarnings")
```


```{r library}
library(earlywarnings)
```
```{r}
library(xts)
library(lubridate)
```


```{r circulation-data-set}
data(circulation)
circ <- xts(circulation$x, order.by = as.Date("2016-01-01") + hours(circulation$time))
str(circ)
```


```{r detrending}
plot(circ)
plot(circulation$time, circulation$x, type="l", xlab="time")
lines(sm <- ksmooth(circulation$time, circulation$x, "normal", bandwidth=50), col="red")
```

```{r residuals}
residuals <- circ - sm$y
residuals <- circulation - sm$y
plot(residuals, type="h", ylab="residuals", xlab="time")
abline(h=0)
```

```{r moving-window}
winsize=50
mw <- round(length(circulation$x) * winsize/100)
```


```{r standard-deviation}
mvSD <- rollapply(residuals, width = mw, FUN = sd, fill=NA, align="right")
plot(mvSD)
```

```{r acor-function}
acor <- function(x){
  cor(x[-1], x[-length(x)])
}
```

```{r acf, eval=F}
str(residuals)
acor(residuals)
mvACF <- rollapply(residuals, width = mw, FUN = acor, fill=NA, align="right")
mvACF <- rollapply(residuals, width = mw, FUN = function(x) acf(coredata(x), na.action=na.omit, lag.max = 1, type = c("correlation") , plot=F), fill=NA, align="right")

as.ts(residuals)
residuals <- residuals[1:10] 
rollapply(residuals, width = mw, FUN = function(x) print(na.omit(x)), fill=NA, align="right")
for(i in 1:length(residuals)){
  acf(residuals[i:(i+mw)], lag.max = 1, type = c("correlation") , plot=F)
}
plot(mvACF)
plot(circulation$time, mvACF, type="l")
```

```{r return-rate, eval=F}
nARR <- rollapply(residuals, width = mw, 
                   FUN = function(x) ar.ols(x, aic = FALSE, order.max = 1, 
      dmean = FALSE, intercept = FALSE)$ar
                   , fill=NA, align="right")

plot(circulation$time, nARR, type="l")
plot(circulation$time, 1/nARR, type="l", ylab="return rate")
```




# Working with paleoclimatic data

We will us data from the National Oceanic and Atmospheric Administration (NOAA).
NOAA's former three data centers---the National Climatic Data Center (NCDC), the National Geophysical Data Center (NGDC), and the National Oceanographic Data Center (NODC), which includes the National Coastal Data Development Center (NCDDC)—have merged into the National Centers for Environmental Information (NCEI).
You will reach all of them over their central domain [ncei.noaa.gov](https://www.ncei.noaa.gov), but they retain their old domains at first. Therefore you will for example still find all climate informations at NCDC's website [ncdc.noaa.gov](http://ncdc.noaa.gov). Paleoclimatic datasets (aka past climate data retained from boreholes) you find under the url [http://www.ncdc.noaa.gov/data-access/paleoclimatology-data/datasets](http://www.ncdc.noaa.gov/data-access/paleoclimatology-data/datasets).


Following two publications ([Livina & Lenton (2010)](http://www.clim-past.net/6/77/2010/cp-6-77-2010.html) and [Lenton et al (2012)](http://www.clim-past.net/8/1127/2012/cp-8-1127-2012.html)) cited at the [homepage of the early warnings signals toolbox ](http://www.early-warning-signals.org/case-studies/paleoclimatic-transitions/) we will use two data sets availbable from NCDC:

The [first dataset](http://www.ncdc.noaa.gov/paleo/study/2496), 
produced by Rasmussen et. al (2007)[^Rasmussen], is the one used in the above mentioned papers. It contains oxygen and calcium Isotope measurements from an ice core taken in Greenland dating way back to more than 30,000 years before present.

[^Rasmussen]:S.O. Rasmussen, I.K. Seierstad, K.K. Andersen, M. Bigler, D. Dahl-Jensen, and S.J. Johnsen. 2007. Synchronization of the NGRIP, GRIP, and GISP2 ice cores across MIS 2 and palaeoclimatic implications. Accepted, Quat. Sci. Rev., INTIMATE special issue, 2007

The [second dataset](https://www.ncdc.noaa.gov/paleo/study/2494), produced by 
Vinther at al (2006)[^Vinther], showing how the oxygen isotope data developed in the Holocene until today.

[^Vinther]:B.M. Vinther, H.B. Clausen, S.J. Johnsen, S.O. Rasmussen, K.K. Andersen, S.L. Buchardt, D. Dahl-Jensen, I.K. Seierstad, M.-L. Siggaard-Andersen, J.P. Steffensen, A.M. Svensson, J. Olsen, and J. Heinemeier. 2006. A synchronized dating of three Greenland ice cores throughout the Holocene. Journ. Geophys. Res., vol. 111, D13102.

## Download the Rasmussen

The Rasmussen dataset can be downloaded from this url:

```{r download-Rasmussen}
RasmussenURL <- "http://www1.ncdc.noaa.gov/pub/data/paleo/icecore/greenland/summit/ngrip/gicc05-calcium-20yr.txt"
download.file(RasmussenURL, basename(RasmussenURL))
```

In the head of the file we read:
```{txt}

DESCRIPTION: 
Calcium concentration and d18O isotope ratios from the Greenland NGRIP, 
GRIP, and GISP2 ice cores on the GICC05 time scale. The period covered 
is 11.6-32.44 ka b2k (years before A.D. 2000).

Calcium concentration and d18O isotope ratios from the Greenland NGRIP, 
GRIP, and GISP2 ice cores resampled to 20 year resolution and presented 
on the GICC05 time scale. The period covered is 11.6-32.44 ka b2k 
(years before A.D. 2000).

[...]

The NGRIP1 core reaches down to a depth of 1372 m. The NGRIP2 core 
(drilled 20 meters away from the NGRIP1 core) reaches from surface 
to bedrock, but high-resolution measurements have only been carried 
out downwards from 1346 m. All depth reported here are NGRIP2 depths. 

[...]


COLUMNS IN THE DATA FILE
 1 : Age (years b2k)
 2 : Maximum counting error (years)
 3 : NGRIP2 depth (m)
 4 : NGRIP d18O (permille)
 5 : NGRIP [Ca^2+] (ppb)
 6 : GRIP depth (m)
 7 : GRIP d18O (permille)
 8 : GRIP [Ca^2+] (ppb)
 9 : GISP2 depth (m)
10 : GISP2 d18O (per mille)
11 : GISP2 [Ca^2+] (ppb)

11620     98     1488.1     -36.41    17.08    1619.17   -37.14    10.91    1673.23   -36.73      8.7
11640     98    1489.26     -36.57    16.41    1620.56    -36.6     9.81    1674.52   -36.44    14.86
11660     98    1490.44     -36.68    22.68    1621.96   -36.81    11.71    1675.83   -36.01    17.75

```



```{r read-Rasmussen, tidy=F}
colnames <- c("Age (years b2k)",
"Maximum counting error (years)",
"NGRIP2 depth (m)",
"NGRIP d18O (permille)",
"NGRIP [Ca^2+] (ppb)",
"GRIP depth (m)",
"GRIP d18O (permille)",
"GRIP [Ca^2+] (ppb)",
"GISP2 depth (m)",
"GISP2 d18O (per mille)",
"GISP2 [Ca^2+] (ppb)")
  
Rasmussen <- read.table(basename(RasmussenURL), header = F, skip= 199, col.names = colnames, na.strings = "NaN")
```

## Download the Vinther data set

We can download the data set by Vinther from this url:

```{r download-vinther}
VintherURL <- "http://www1.ncdc.noaa.gov/pub/data/paleo/icecore/greenland/summit/ngrip/gicc05-holocene-20yr.txt"
download.file(VintherURL, basename(VintherURL))

```

In the head of the data file we read:

```{txt}
DESCRIPTION: 
Greenland Ice Core Chronology 2005 (GICC05) Holocene section.
Entire Holocene in 20-year resolution and d18O from NGRIP1, NGRIP2, GRIP, and DYE-3.
All ages are given in years b2k (years relative to A.D. 2000).
The concept of maximum counting error is introduced in Rasmussen et al. 
(Journ. Geophys. Res, 2006)

[...]


COLUMNS IN THE DATA FILE:
 1 : Age (years b2k)
 2 : NGRIP1 depth (m)
 3 : NGRIP1 d18O (permille)
 4 : NGRIP2 depth (m)
 5 : NGRIP2 d18O (permille)
 6 : GRIP depth (m)
 7 : GRIP d18O (permille)
 8 : DYE-3 depth (m)
 9 : DYE-3 d18O (permille) SEE NOTE (*) ABOVE
10 : Maximum counting error (years)


20         7.35    -35.16      NaN       NaN       5.43      NaN          0      NaN      1
40        14.15    -35.82      NaN       NaN       13.6    -35.13     20.21    -27.36     1
60         20.4    -35.11      NaN       NaN      21.43    -35.55     37.24    -27.06     1

```

Therefore we read it like this:

```{r read-Vinther, tidy=F}
colnames <- c(
"Age (years b2k)",
"NGRIP1 depth (m)",
"NGRIP1 d18O (permille)",
"NGRIP2 depth (m)",
"NGRIP2 d18O (permille)",
"GRIP depth (m)",
"GRIP d18O (permille)",
"DYE-3 depth (m)",
"DYE-3 d18O (permille)",
"Maximum counting error (years)"
  )
Vinther <- read.table(basename(VintherURL), header = F, skip= 109, 
                      col.names = colnames, na.strings = "NaN")
```


## Search for early warning signals in the paleo data

First we create a zoo timeseries from the Rasmussen and the Vinther GRIP $$\delta^{18}$$O variable. They overlap by 5 values so we cut off some from the Rasmussen data set.

```{r, zoo-timeseries}
paleoz <- zoo(c(Vinther$GRIP.d18O[-1], Rasmussen$GRIP.d18O[-(1:5)]), order.by = c(Vinther$Age[-1], Rasmussen$Age[-(1:5)]))
```

Then we will fill in the gaps that this timeseries has  by interpolating with the `na.approx()` function of the `zoo` package.

```{r fill-gaps}
#paleoz <- na.approx(paleoz)
```

The next task is to smooth out the high frequency oscillations and produce a trendline for the series. We choose a bandwith of 558 like they did in the [Lenton et al (2012)](http://www.clim-past.net/8/1127/2012/cp-8-1127-2012.html) paper:

```{r smooth}
sm <- ksmooth(c(Vinther$Age[-1], Rasmussen$Age[-(1:5)]), paleoz, "normal", bandwidth=558)
```

All this together plots like this:

```{r plot-all}
par(mar=c(4,4,.1,2), pch=20, mgp=c(2,1,0))
#xlim=c(22000, 12000)
xlim=c(32440, 0)
#plot(Vinther$Age, Vinther$GRIP.d18O, type="l", xlim=xlim, ylim=c(-45, -32), col="lightblue", xlab="Age")
plot(paleoz, xlim=xlim)
abline(v=c(12700, 14700, 11700))
lines(sm, col="red")
```

We then calculate the residuals representing the detrended short-term oscillations:

```{r Rasmussen-residuals}
residuals <- paleoz - sm$y
plot(residuals, type="h", xlim=xlim)
```


```{r, rolling-window}
window <-250
mvSD <- rollapply(residuals, width = window, FUN = var, na.rm=T, fill=NA, partial=F, align="left")
mvACF <- rollapply(residuals, width = window, FUN = acor, fill=NA, partial=F, align="left")
nARR <- rollapply(residuals[-(1:42)], width = window, 
                  FUN = function(x) ar.ols(x, 
    aic = FALSE, order.max = 1, dmean = FALSE, intercept = FALSE)$ar, 
    fill = NA, align = "right")


plot(mvSD, type="l", xlim=xlim, col="red")
par(new=T)
plot(mvACF, type="l", xlim=xlim, col="green", ann=F, axes=F)
axis(4, col="green")
mtext("mvACF", 4, line=3, las=3, col="green")
par(new=T)
plot(nARR, xlim=xlim, type = "l", col="blue")
par(new=T)
plot(1/nARR, xlim=xlim, type = "l", ylab = "return rate", col="orange")

```

<!----------------------------------------------------------------->

```{r, eval=F, echo=F}
library(earlywarnings)
pal <- data.frame(rev(-c(Vinther$Age[-1], Rasmussen$Age[-(1:5)])), rev(paleoz))
out=generic_ews(pal,winsize=50,detrending='gaussian',
bandwidth=5,logtransform=FALSE,interpolate=FALSE)


plot(sd~timeindex, data=out)
```

```{r, eval=F, echo=F}
out=generic_ews(circulation,winsize=50,detrending='gaussian',
bandwidth=5,logtransform=FALSE,interpolate=FALSE)
plot(sd~timeindex, data=out)
```

```{r, eval=F, echo=F}
X = c(rnorm(1000, mean = 0), rnorm(1000, mean = -2), rnorm(1000, mean = 2)); 
param = seq(0,5,length=3000); 
res <- movpotential_ews(X, param)

X = c(rnorm(1000, mean = 0), rnorm(1000, mean = -2), rnorm(1000, mean = 2)); 
param = seq(0,5,length=3000); 
res <- movpotential_ews(X, param)

data(foldbif)
bdstest_ews(circulation,ARMAoptim=FALSE,ARMAorder=c(1,0),embdim=3,epsilon=0.5, 
boots=200,logtransform=FALSE,interpolate=FALSE)

output<-ddjnonparam_ews(circulation,bandwidth=0.6,na=500,
logtransform=TRUE,interpolate=FALSE)

sensitivity_ews(circulation,indicator='sd',detrending='gaussian',
incrwinsize=25,incrbandwidth=20)

output=surrogates_ews(foldbif,indicator='sd',winsize=50,detrending='gaussian',
bandwidth=10,boots=200,logtransform=FALSE,interpolate=FALSE)

```


