---
layout: post
title: "Summary stats with dplyr"
excerpt: "calculate monthly mean"
category: tutorial
language: English
author: "Nanu Frechen"
datasource: <a href="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/1_minute">DWD</a>
technique: <a href="https://readr.tidyverse.org/">readr</a>, <a href="https://dplyr.tidyverse.org/">dplyr</a>, <a href="https://ggplot2.tidyverse.org/">ggplot2</a>, <a href="https://lubridate.tidyverse.org/">lubridate</a>
---
  
* auto-gen TOC:
{:toc}

Here I show how to do some simple summary statistics with the [dplyr package](https://dplyr.tidyverse.org/).


We will be using the followint packages:  
  
```{r setup, echo=T, message = F, error = F, warning = F}
library(lubridate)
library(dplyr)
library(readr)
library(ggplot2)
```

```{r ggplot theme, echo=F, include=F, cache=F}
library(knitr)
opts_chunk$set(message = F, error = F, warning = F, cache = T, autodep=T, dev="png", fig.width=9, fig.height=6, dev.args=list(bg="transparent"), dpi=300)

theme_set(theme_linedraw() +
            theme(
              panel.grid.major = element_line(size=0.2, color="gray80"),
              panel.grid.minor= element_line(size=0.1, color="gray80"),
              text = element_text(size=14, family = "Open Sans"),
              plot.title = element_text(hjust = 0.5, size=14),
              plot.subtitle = element_text(hjust=0.5, size=12),
              plot.caption = element_text(size=8),
              plot.margin = unit(c(4,5.5,5.5,5.5), "pt"),
              strip.background = element_blank(),
              strip.text = element_text(colour = "black"),
              panel.background = element_rect(fill = "transparent"), # bg of the panel
              plot.background = element_rect(fill = "transparent", color = NA), 
              legend.background = element_rect(fill="transparent", color=NA),
              legend.key = element_rect(fill="transparent", color=NA),
              legend.title = element_text(size=12),
              legend.text = element_text(size=12),
              legend.position="right",
              legend.direction="vertical",
              # axis.title.y=element_text(angle=90, vjust=15)
            )
)

scale_x_month <- function(breaks, labels, ...){
    scale_x_continuous(breaks=c(yday(as.Date("2020-01-01")+months(0:11)), 365), labels=month.abb[c(1:12,1)], ...)
}
```

# get DWD data

- [station list](https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/KL_Tageswerte_Beschreibung_Stationen.txt)
- [variables description](https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/DESCRIPTION_obsgermany_climate_daily_kl_historical_en.pdf)


Files can be downloaded and unzipped by hand from [https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical](https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical). Or you can use the following script:

```{r download data}
data_dir = "data/dwd"
dir.create(data_dir)

urls <- c(Cottbus="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/tageswerte_KL_00880_18870101_20231231_hist.zip",
          Warnemünde="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/tageswerte_KL_04271_19470101_20231231_hist.zip",
          #Freiberg="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/tageswerte_KL_01441_19450701_19930430_hist.zip",
          #Stuttgart="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/tageswerte_KL_04927_19490801_19840731_hist.zip",
          "Garmisch-Patenkirchen"="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/tageswerte_KL_01550_19360101_20231231_hist.zip",
          Zugspitze="https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/tageswerte_KL_05792_19000801_20231231_hist.zip")

for(url in urls){
  destfile = file.path(data_dir, basename(url))
  if(!file.exists(destfile)){
    download.file(url, destfile)
    unzip(destfile, exdir = data_dir)
  }
}
```



# read data

We use the [readr package](https://readr.tidyverse.org/) to read the csv files:

```{r read data}
data_files <- list.files(data_dir, pattern="produkt_klima_tag", full.names = T)

data <- read_delim(data_files, delim = ";", na = "-999", trim_ws = T) 
```
We use `list.files` to list all files starting with "produkt_klima_tag" in the data directory. The `read_delim` function reads all files into a single R object. Stations can be identified by their STATIONS_ID in the dataset.


See what's in the dataset:

```{r show data}
data
```


# convert data

Convert date:

```{r convert date}
data <- data %>%
  mutate(MESS_DATUM = as.POSIXct(as.character(MESS_DATUM), format="%Y%m%d")) %>%
  mutate(STATIONS_ID = as.factor(STATIONS_ID))
```

The `%>%` notation and `mutate` function comes from the [dplyr package](https://dplyr.tidyverse.org//).

Calculate yday and month:

```{r calculate yday and month}
data <- data %>%
  mutate(yday = yday(MESS_DATUM)) %>%
  mutate(month = month(MESS_DATUM))
```

The `yday` and `month` functions come from the [lubridate package](https://lubridate.tidyverse.org/)

Add station names:

```{r add statin names}
stations <- tribble(
  ~STATION_NAME, ~STATIONS_ID,
  "Cottbus", "880",
  "Warnemünde", "4271",
  "Garmisch-Patenkirchen", "1550",
  "Zugspitze", "5792"
)

data <- data %>% left_join(stations)
```


# plot some data

We plot data with the [ggplot2 package](https://ggplot2.tidyverse.org/):

```{r plot temperature data}
data %>%
  ggplot() +
  aes(x=yday, y=TMK) +
  geom_point(size=0.1, alpha=0.1) +
  facet_wrap(STATION_NAME~.)
```
`facet_wrap` creates a single graph for each station.

# calculate mean for month

```{r calc mothly average}
longterm_average_month <- data %>%
  group_by(STATION_NAME, month) %>%
  summarise("temp longterm average" = mean(TMK, na.rm=T), "rain longterm average" = mean(RSK, na.rm=T))
```

With `group_by` and `summarise` we again use functions from the [dplyr package](https://dplyr.tidyverse.org//). The procedure is to form multiple subsets of the dataset with `group_by` and then calculate the summary statistics for those subsets. We group by STATION_NAME and month.

In the next plot we put all stations into one graph:

```{r plot temp monthly average, fig.height=4}
longterm_average_month %>%
  ggplot() +
  aes(x=month, y=`temp longterm average`, col=STATION_NAME, group=STATION_NAME) +
  geom_line()
```



# calculate mean for yday

```{r calc yday average}
longterm_average_yday <- data %>%
  group_by(STATION_NAME, yday) %>%
  summarise("temp longterm average" = mean(TMK, na.rm=T), "rain longterm average" = mean(RSK, na.rm=T))
```


```{r plot yday average with scatterplot}
data %>%
  ggplot() +
  aes(x=yday, y=TMK) +
  geom_point(size=0.1, alpha=0.1) +
  geom_line(aes(x=yday, y=`temp longterm average`), data=longterm_average_yday, col="red") +
  facet_wrap(STATION_NAME~.)
```
# precipitation

```{r rain scatterplot}
data %>%
  ggplot() +
  aes(x=yday, y=RSK) +
  scale_y_continuous(trans="log10") +
  geom_point(size=0.1, alpha=0.1) +
  facet_wrap(STATION_NAME~.)
```

```{r rain monthly average plot}
longterm_average_month %>%
  ggplot() +
  aes(x=month, y=`rain longterm average`, col=STATION_NAME, group=STATION_NAME) +
  #scale_y_continuous(trans="log10") +
  geom_line()
```

